Is the direction of Granger causal influence the same as the direction of information flow?

### 2015-07-07 paper.tex
Make misc. edits; edit numerical results section.

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Some thought experiments on the applicability of Granger causality and Directed Information in inferring direction of information flows}
\author{Praveen Venkatesh and Pulkit Grover\\Carnegie Mellon University}

@@

\begin{abstract}
	Granger causality is an established measure of the ``causal influence'' that one stochastic process has on another. Along with its more recent generalization -- Directed Information -- Granger Causality has been used extensively in neuroscience, and in complex interconnected systems in general, to infer causal influences. Of late, many works have begun to interpret the direction of causal influence as the direction of ``information flow''. We ask: is the direction of causal influence, as predicted by Granger Causality and Directed Information always the same as the direction of information flow? We test whether these measures correctly predict the direction of information flow by using two simple theoretical experiments, in which the true direction of information flow is known by design. The experiments are based on a communication system with a feedback channel, upon which a strategy inspired by the work of Schalkwijk and Kailath is employed. We show that in these experiments, the direction of information flow can be opposite to that which is inferred using Granger causality and Directed Information. Thus, while it might be reasonable to infer the direction of causal influence using these techniques, one needs to exercise care in interpreting the direction of causal influence as the direction of information flow.
\end{abstract}

@@

We note here that Granger's original analysis does not compare this metric on forward and reverse links, however, this appears to be commonly done in practice. {\color{red} TODO: Cite Tsachy's paper here.}

@@

It is important to note that the technical objections listed above are all concerned with \emph{measurement}. They indicate that incorrect Granger Causal influences may be computed if there is some deficiency in the measurement procedure. These objections can be resolved by taking better measurements (by sampling more nodes, with higher signal-to-noise ratio, etc.).

Our objection, on the other hand, is much more fundamental. \emph{Even if} the measurements are made with infinite accuracy, and the regression coefficients associated with computing Granger Causality are precisely estimated, and the Granger Causality metric is perfectly computed (as is the case in our counter-examples), Granger Causal influence would \emph{still} not give the correct direction of information flow. To our knowledge, this is a completely novel argument against Granger Causality, and one that is far more serious than previous objections, at least as far as its usefulness in understanding brain function is concerned.

@@

\subsection{Discussion}
\begin{enumerate}
	\item \color{red} Some of this misconception might have arisen because of Massey's directed info conservation - we expect that dir info in the forward dirn goes up if the other one goes down - but this is a misnomer. (Do we still need this? I am increasingly of the opinion that directed info / GC capture nothing about the message)
\end{enumerate}

%Notes:

%You only seem to be assuming that the variance is $\sigma_\theta^2$. Nothing about whether it is a uniform random variable or something derived from quantizing a uniform random variable, or a Gaussian one.

@@

\subsection{Granger causality for the noiseless feedback case}

@@

\subsection{Directed Information for the noiseless feedback case}

@@

\section{Numerical results}
\label{sec:numerical-results}
In the noiseless feedback case, the directed information on the forward link is finite, while that on the feedback link is infinite. However, the situation is more interesting when there is noise in the feedback link, as is illustrated in Fig.~\ref{fig:numerical} through numerical calculations of expressions in the last section.

@@

>>>>

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Is the direction of Granger causal influence the same as the direction of information flow?}
\author{Praveen Venkatesh\\vpraveen@cmu.edu\\Carnegie Mellon University}
\author{Pulkit Grover\\pulkit@cmu.edu\\Carnegie Mellon University}

@@

\begin{abstract}
	Granger causality is an established measure of the ``causal influence'' that one stochastic process has on another. Along with its more recent generalization -- Directed Information -- Granger Causality has been used extensively in neuroscience, and in complex interconnected systems in general, to infer causal influences. Of late, many works have begun to interpret the direction of causal influence as the direction of ``information flow''. We ask: is the direction of causal influence, as predicted by Granger Causality and Directed Information always the same as the direction of information flow? We test whether these measures correctly predict the direction of information flow by using two simple theoretical experiments, in which the true direction of information flow (the ``ground truth'') is known by design. The experiments are based on a communication system with a feedback channel, upon which a strategy inspired by the work of Schalkwijk and Kailath is employed. We show that in these experiments, the direction of information flow can be opposite to that which is inferred using Granger causality and Directed Information. Thus, while it might be reasonable to infer the direction of causal influence using these techniques, one needs to exercise care in interpreting the direction of causal influence as the direction of information flow.
\end{abstract}

@@

We note here that Granger's original analysis does not compare this metric on forward and reverse links, however, this is commonly done in practice even in areas beyond neuroscience (eg.~\cite{Brovelli2004BetaOscillations,jiao2013universal}).

@@

It is important to note that the technical objections listed above are all deficiencies in or limitations of \emph{measurement}. They indicate that incorrect Granger Causal influences may be computed if there is some deficiency or limitation in the measurement procedure. These objections can be resolved by taking better measurements (by sampling more nodes, with higher signal-to-noise ratio, etc.).

Our objection, on the other hand, is more fundamental. \emph{Even if} the measurements are made with infinite accuracy, and the regression coefficients associated with computing Granger Causality are precisely estimated, and the Granger Causality metric is perfectly computed (as is the case in our counter-examples), Granger Causal influence may \emph{still} not yield the correct direction of information flow. To our knowledge, the argument that Granger causal influence can be opposite to the direction of information flow is a novel one. We believe that this argument is much more serious than previous objections, at least in the context of determining the directions of information flow in neuroscientific experiments, towards understanding the computational functions of brain regions.

@@

%Notes:
%You only seem to be assuming that the variance is $\sigma_\theta^2$. Nothing about whether it is a uniform random variable or something derived from quantizing a uniform random variable, or a Gaussian one.

@@

\subsection{Granger causality for the noiseless feedback case}
\label{sec:gc-noiseless}

@@

\subsection{Directed Information for the noiseless feedback case}
\label{sec:dir-info-noiseless}

@@

\section{Numerical results demonstrating that Granger Causal influence can be opposite in direction to information flow}
\label{sec:numerical-results}
In the noiseless feedback case, the directed information on the forward link is finite, while that on the feedback link is infinite (refer sections~\ref{sec:gc-noiseless} and~\ref{sec:dir-info-noiseless}). However, for completeness, we examine the case when noise is present in the feedback link, as is illustrated in Fig.~\ref{fig:numerical}, through numerical calculation of expressions in the last section. For cases where the noise variance of the feedback link ($\sigma_R^2$) is moderately smaller than feedforward noise variance ($\sigma_N^2$), we observe that the Directed Information in the direction of the reverse link can dominate that in the direction of the forward link. With sufficiently many (albeit sometimes a large number of) iterations, the Directed Information in forward direction starts to dominate. However, the point at which this happens depends on the (often unknown) ratio of noise in these links.

@@

% TODO: Include in final version:
% Some of this misconception might have arisen because of Massey's directed info conservation - we expect that dir info in the forward dirn goes up if the other one goes down - but this is a misnomer. (Do we still need this? I am increasingly of the opinion that directed info / GC capture nothing about the message)
